Towards a standardized evaluation of imputation methodology

### 2022-07-26 Manuscript/Towards_a_standardized_evaluation_of_imputation_methodology.tex
Add TODOs for revision

\documentclass[bimj,fleqn]{w-art}
\usepackage{times}
\usepackage{w-thm}
\usepackage[authoryear]{natbib}
% \setlength{\bibsep}{2pt}
% \setlength{\bibhang}{2em}
% \newcommand{\J}{J\"{o}reskog}
% \newcommand{\So}{S\"{o}rbom}
% \newcommand{\bcx}{{\bf X}}
% \newcommand{\bcy}{{\bf Y}}
% \newcommand{\bcz}{{\bf Z}}
% \newcommand{\bcu}{{\bf U}}
% \newcommand{\bcv}{{\bf V}}
% \newcommand{\bcw}{{\bf W}}
% \newcommand{\bci}{{\bf I}}
% \newcommand{\bch}{{\bf H}}
% \newcommand{\bcb}{{\bf B}}
% \newcommand{\bcr}{{\bf R}}
% \newcommand{\bcm}{{\bf M}}
% \newcommand{\bcf}{{\bf F}}
% \newcommand{\bcg}{{\bf G}}
% \newcommand{\bcs}{{\bf S}}
% \newcommand{\bca}{{\bf A}}
% \newcommand{\bcd}{{\bf D}}
% \newcommand{\bcc}{{\bf C}}
% \newcommand{\bce}{{\bf E}}
% \newcommand{\ba}{{\bf a}}
% \newcommand{\bb}{{\bf b}}
% \newcommand{\bc}{{\bf c}}
% \newcommand{\bd}{{\bf d}}
% \newcommand{\bx}{{\bf x}}
% \newcommand{\by}{{\bf y}}
% \newcommand{\bz}{{\bf z}}
% \newcommand{\bu}{{\bf u}}
% \newcommand{\bv}{{\bf v}}
% \newcommand{\bh}{{\bf h}}
% \newcommand{\bl}{{\bf l}}
% \newcommand{\be}{{\bf e}}
% \newcommand{\br}{{\bf r}}
% \newcommand{\bw}{{\bf w}}
% \newcommand{\de}{\stackrel{D}{=}}
% \newcommand{\bt}{\bigtriangleup}
% \newcommand{\bfequiv}{\mbox{\boldmath $\equiv$}}
% \newcommand{\bmu}{\mbox{\boldmath $\mu$}}
% \newcommand{\bnu}{\mbox{\boldmath $\nu$}}
% \newcommand{\bxi}{\mbox{\boldmath $\xi$}}
% \newcommand{\btau}{\mbox{\boldmath $\tau$}}
% \newcommand{\bgamma}{\mbox{\boldmath $\Gamma$}}
% \newcommand{\bphi}{\mbox{\boldmath $\Phi$}}
% \newcommand{\bfphi}{\mbox{\boldmath $\varphi$}}
% \newcommand{\bfeta}{\mbox{\boldmath $\eta$}}
% \newcommand{\bpi}{\mbox{\boldmath $\Pi$}}
% \newcommand{\bequiv}{\mbox{\boldmath $\equiv$}}
% \newcommand{\bvarepsilon}{\mbox{\boldmath $\varepsilon$}}
% \newcommand{\btriangle}{\mbox{\boldmath $\triangle$}}
% \newcommand{\bdelta}{\mbox{\boldmath $\Delta$}}
% \newcommand{\beps}{\mbox{\boldmath $\epsilon$}}
% \newcommand{\btheta}{\mbox{\boldmath $\theta$}}
% \newcommand{\balpha}{\mbox{\boldmath $\alpha$}}
% \newcommand{\bsphi}{\mbox{\boldmath $\varphi$}}
% \newcommand{\bsig}{\mbox{\boldmath $\sigma$}}
% \newcommand{\bfpsi}{\mbox{\boldmath $\psi$}}
% \newcommand{\bfdelta}{\mbox{\boldmath $\delta$}}
% \newcommand{\bsigma}{{\bf \Sigma}}
% \newcommand{\bzero}{{\bf 0}}
% \newcommand{\bpsi}{\mbox{\boldmath $\Psi$}}
% \newcommand{\bep}{\mbox{\boldmath $\epsilon$}}
% \newcommand{\bomega}{\mbox{\boldmath $\Omega$}}
% \newcommand{\bfomega}{\mbox{\boldmath $\omega$}}
% \newcommand{\blambda}{\mbox{\boldmath $\Lambda$}}
% \newcommand{\bflambda}{\mbox{\boldmath $\lambda$}}
% \newcommand{\bfsigma}{\mbox{\boldmath $\sigma$}}
% \newcommand{\bfpi}{{\mbox{\boldmath $\pi$}}}
% \newcommand{\bupsilon}{\mbox{\boldmath $\upsilon$}}
% \newcommand{\obs}{{\rm obs}}
% \newcommand{\mis}{{\rm mis}}
% \theoremstyle{plain}
% \newtheorem{criterion}{Criterion}
% \theoremstyle{definition}
% \newtheorem{condition}[theorem]{Condition}
% \usepackage[]{graphicx}
% \chardef\bslash=`\\ % p. 424, TeXbook
% \newcommand{\ntt}{\normalfont\ttfamily}
% \newcommand{\cn}[1]{{\protect\ntt\bslash#1}}
% \newcommand{\pkg}[1]{{\protect\ntt#1}}
% \let\fn\pkg
% \let\env\pkg
% \let\opt\pkg
% \hfuzz1pc % Don't bother to report overfull boxes if overage is < 1pc
% \newcommand{\envert}[1]{\left\lvert#1\right\rvert}
% \let\abs=\envert

@@

\begin{abstract}
Developing new imputation methodology has become a very active field. Unfortunately, there is no consensus on how to perform simulation studies to evaluate the properties of imputation methods. In this paper, we propose a move towards a standardized evaluation of imputation methodology. To demonstrate the need for standardization, we highlight a set of potential pitfalls that bring forth a chain of potential problems in the objective assessment of the performance of imputation routines. This may lead to sub-optimal use of  imputation in practice. Additionally, we suggest a course of action for simulating and evaluating missing data problems. [TODO: explain how the standardisation leads to more neutrality, but that this is just a set of elements to consider/report, not a complete 'cookbook'.]
\end{abstract}

@@

[TODO: Narrowing the scope of the manuscript to the evaluation of methodology for statistical inference, not prediction or causal inference (although inferential validity encompasses predictive validity and may thus be generalized?). Emphasizing that the ideal evaluation of imputation methodology depends on the aim of the specific study (i.e. how the evaluated methods are used in practice).]

Imputation is a state-of-the-art technique for drawing valid conclusions from incomplete data. The technique has earned a permanent spot in research and policymaking, demonstrated e.g. by the detailed manual created by the National Research Council \citep*{little2012prevention}. Although top-down enforcement of valid ways to handle missing data is not yet very pronounced, an increasing amount of researchers and data scientists are embracing imputation techniques. After all, the principle of imputation is very intuitive.

The idea behind imputation is to impute (fill in) missing values, to obtain a valid estimate of what could have been. A popular variant is multiple imputation \citep{rubi76}, whereby each missing value is imputed multiple times. The completed data that are thus obtained can be analyzed by standard techniques and, in the case of multiple imputation, the analysis results can be combined into a single inference using Rubin's rules \citep[][p. 76]{rubi87}. In contrast to ad hoc methods for dealing with missing values (e.g. list-wise deletion, mean imputation, regression imputation, last observation carried forward, indicator method), multiple imputation properly takes into account the sources of uncertainty that are related to the missingness problem.

The quality of a solution obtained by imputation depends on the statistical properties of the incomplete data and the degree to which an imputation procedure is able to capture these properties when modeling missing values. In general, it holds that modeling missing data becomes more challenging when the amount of missingness increases. However, when (strong) relations in the data are present, the observed parts can hold great predictive power for the models that estimate the missingness. In that case, imputation would be substantially more efficient than the ubiquitous complete case analysis.

When evaluating the statistical properties (and thereby the practical applicability) of imputation methodology, researchers most often make use of simulation studies. In such studies a complete dataset is usually generated from a statistical model, another model is used to induce missingness, and a set of evaluation criteria is postulated to evaluate the performance of one or more missing data methods. However, no gold standard has been established to evaluate imputation routines and, as a result, the validity and aim of the simulation setups may differ tremendously from one developer to another. Especially with novel imputation methods being propagated from the fields of machine learning and artificial intelligence, the differences may become more pronounced. Although these promising new methods seem to yield even sharper imputations than now-standard (semi-)parametric imputation methods, the comparison may not be fair due to the simulation setup.

The purpose of this paper is threefold: First, to raise some concerns with respect to evaluating imputation methodology. These concerns stem from careful consideration with fellow `imputers' and from encounters as a reviewer for statistical journals. Second, to provide imputation methodologists with a suggested course of action when simulating missing data problems. This suggested approach should identify common ground but is in no way intended as an absolute solution. This identifies the third purpose of this paper: discussion. We hope to elicit critical thinking regarding the problems at hand. We are all convinced that our methodology has some merit. But for sake of progress, it would be much more advantageous if the aim of our evaluations would go beyond \textit{proving the point} and would legitimately consider the statistical properties. [TODO: explain how the standardisation leads to more neutrality, but that this is just a set of elements to consider/report, not a complete 'cookbook'.]

Standardizing the evaluation of imputation methodology requires simulators to consider a couple of key aspects in their simulation workflows. Table \ref{table:steps} outlines some suggested steps to adopt. Please note that there is--of course--no `silver bullet' and that we do not claim to present a universally applicable approach. Our suggestions are the accumulated result of scientific literature, research experience, and discussion. For an excellent overview of general best practices for method evaluation by means of simulation, see \citet{morr18}. We recommend adhering to their proposed ADEMP structure (aims, data-generating mechanisms, estimands, methods, performance measures) for planning and reporting simulation studies, but highlight and add some specific recommendations for the evaluation of imputation methodology.

@@

[TODO: Split up the problems with the simulation setup/evaluation and the problems with the interpretation/extrapolation from these simulation studies. See e.g. \citet{gree17} on cognitive problem of generelization versus methodological problems.]

As of today, there is no consensus on how to perform simulation studies to evaluate the properties of imputation methods. Typically, the developer of a new imputation routine does some tests by simulations, but these tests differ across developers. This brings forth a chain of potential problems in the objective assessment of imputation method performance, within and across studies, which may lead to sub-optimal use of imputation in practice. To demonstrate the broad impact of these problems, we subdivide the problems into the following three distinct categories: problems with data generation, problems with missingness generation, and problems with performance evaluation. We further detail the impact each of these problems may have on the validity of evaluations.

@@

\subsection{Study scope}

[Problem: parameters unclear, extrapolation beyond scope of simulations, misinterpretation \citep[see "investigator bias"][]{gree17}.]

Before setting up a simulation study, the simulator should clearly define the scope of their evaluations. A simulation study aimed at imputation method comparison will typically have a different design than one aimed at establishing the inferential validity of a single (novel) method. The choice of imputation method(s) under evaluation is naturally intertwined with the simulation aims and setup.

One specific simulation study aim (that we have thus far ignored) is to evaluate the predictive performance of pairs of imputation methods and estimation methods in empirical, incomplete data. These kinds of simulation studies do not start out from a complete data set in which missingness is induced by the simulator. Rather, the evaluations use the observed values of an outcome variable (or target) as the estimand, which is estimated from the incomplete predictor (or feature) space by first imputing the missingness and then applying a prediction method. Pairs of imputation and prediction methods with high predictive accuracy in one or more benchmark data sets are deemed as `good' \citep{liu21}. Note that such a simulation design does not have a `ground truth', so only the comparative performance of methods may be established. We do not recommend this approach if the inferential validity of the imputations is of interest.

Some general choices in the simulation setup to consider at this stage are the simulation design and the number of simulation repetitions. The simulation design should ideally be fully factorial, i.e. varying each simulation condition against all other conditions. There are, for example, known interaction effects between missingness mechanisms and missingness patterns (the validity of the assumed missingness mechanism becomes increasingly important with higher missingness proportions). If there is more than one imputation method under evaluation, the simulation design should apply each method to every incomplete data set. Applying all methods to the same incomplete data set is computationally convenient, and minimizes unnecessary variation, which makes for fairer comparisons. The number of simulation repetitions may be informed by the required level of precision in the simulation study \citep[e.g. as determined from a maximum tolerable level of uncertainty in terms of a performance measure's Monte Carlo error][]{morr18}.

@@

To evaluate the ability of an imputation routine to handle missingness, a form of truth has to be established. Those who perform simulation studies are in the luxury position to establish the truth beforehand by choosing a data-generating mechanism. Data-generating mechanisms define how a complete dataset is obtained at the start of each simulation repetition. There are two general approaches to generating complete data: (i) model-based simulation, in which data are drawn from a known statistical model or probability distribution, such as the multivariate normal distribution; and (ii) design-based simulation, where data are sampled (with replacement) from a sufficiently large observed set, such as official registers.

@@

After deciding upon a data-generating mechanism, another influential--but often overlooked--decision awaits. Simulation studies on missing data methodology have the unique option to exclude sampling variance from their evaluations and only use the missing data generation procedure as the source of Monte Carlo variation. After all, we are interested in the missingness and are not considering the noise induced by the sampling mechanism for evaluation in such studies. Therefore, taking sampling variation into account is neither necessary for obtaining information about a method's ability to handle the missing data problem, nor for objectively comparing methods on their ability to correct for missingness \citep[see for a detailed discussion][]{vink14}. The only required change to simulation setups is to remove sampling variance from performance evaluation. Still, a lot of published work does not consider this option, while it could have sharpened inconclusive or obfuscated results. [TODO: add nuance because this doesn't always hold, see Tims review!]

@@

The simulator should choose their data-generating mechanism(s) in line with the study scope. Model-based data-generating mechanisms have advantages in flexibility and precision, since data are generated from a known statistical model and the true theoretical parameters can be derived. A design-based approach is often used in situations where a probability distribution is not available, or where real-life data structures are of interest.

Estimand(s) or other simulation targets should be defined in the context of the study aim and data-generating mechanism. As a general rule, it would be wise to include one or more multivariate estimands. [TODO: make this part about the study aim and refer to \citet{pete14} about clear and unambiguous estimand, and to focus on this estimand.] It is a realistic expectation for imputation methods to preserve both marginal and conditional distributions with respect to the comparative truth. Imputation methods that fail to do so, should not be considered general-purpose methods (e.g., mean imputation).

The comparative truth of the estimand(s) is determined by the choice of data-generating mechanism and whether sampling variance is included in the simulation setup, see Table \ref{table:dgm_truth}. If sampling variance cannot be omitted from the simulation scheme, multiple samples should be drawn from the data-generating mechanism (i.e., one sample per simulation repetition, a standard simulation setup). If sampling variance is not of interest, a single complete data set can be obtained from the data-generating mechanism (i.e., one sample for all simulation repetitions). This process is computationally convenient because only a single complete data set has to be considered during all of the simulations. We highly value the additional flexibility that such a simulation setup offers. Especially in the case of data transformations, it can be challenging to derive the true parametric references to evaluate against. Simply using a single generated complete set in which missingness is induced, and evaluating against that generated data set avoids a plethora of procedural problems and computational challenges.
When deviating from the standard simulation workflow, conventional pooling rules for multiple imputation \citep[cf.][p. 76-77]{rubi87} do not apply. Instead, alternative pooling rules need to be used \citep{raghunathan2003multiple,vink14}.

@@

[TODO: change this to 'be clear which mechanisms are of interest, how they are generated, if they are suitably generated']

Evaluating imputation methodology requires a missing data problem to be solved. After establishing a comparative truth from a data-generating mechanism, some form of missingness therefore has to be induced. Often, however, reports of simulation studies remain vague about the actual missingness conditions under investigation and, even worse, some authors only report something like:
\begin{quote}
\textit{We generated missing data following a missing at random missingness mechanism.}
\end{quote}
This should be considered unacceptable as claims about the validity of the imputation inference heavily depend on the simulated missingness conditions, such as missingness mechanisms and missingness patterns. Missingness mechanisms describe the relationship between missing entries and observed data values and are generally categorized into MCAR, MAR and MNAR mechanisms \citep[][see Table \ref{table:mech}]{rubi76}. Missingness patterns concern the location of missing entries across incomplete data \citep[][p. 8]{litt20}. Under this definition, there is a row-wise element to the missingness pattern describing which variables are jointly observed, and a column-wise element encompassing the amount of missingness in the data.

>>>>

\documentclass[bimj,fleqn]{w-art}
\usepackage{times}
\usepackage{w-thm}
\usepackage[authoryear]{natbib}
\usepackage{amsfonts}
% \setlength{\bibsep}{2pt}
% \setlength{\bibhang}{2em}
% \newcommand{\J}{J\"{o}reskog}
% \newcommand{\So}{S\"{o}rbom}
% \newcommand{\bcx}{{\bf X}}
% \newcommand{\bcy}{{\bf Y}}
% \newcommand{\bcz}{{\bf Z}}
% \newcommand{\bcu}{{\bf U}}
% \newcommand{\bcv}{{\bf V}}
% \newcommand{\bcw}{{\bf W}}
% \newcommand{\bci}{{\bf I}}
% \newcommand{\bch}{{\bf H}}
% \newcommand{\bcb}{{\bf B}}
% \newcommand{\bcr}{{\bf R}}
% \newcommand{\bcm}{{\bf M}}
% \newcommand{\bcf}{{\bf F}}
% \newcommand{\bcg}{{\bf G}}
% \newcommand{\bcs}{{\bf S}}
% \newcommand{\bca}{{\bf A}}
% \newcommand{\bcd}{{\bf D}}
% \newcommand{\bcc}{{\bf C}}
% \newcommand{\bce}{{\bf E}}
% \newcommand{\ba}{{\bf a}}
% \newcommand{\bb}{{\bf b}}
% \newcommand{\bc}{{\bf c}}
% \newcommand{\bd}{{\bf d}}
% \newcommand{\bx}{{\bf x}}
% \newcommand{\by}{{\bf y}}
% \newcommand{\bz}{{\bf z}}
% \newcommand{\bu}{{\bf u}}
% \newcommand{\bv}{{\bf v}}
% \newcommand{\bh}{{\bf h}}
% \newcommand{\bl}{{\bf l}}
% \newcommand{\be}{{\bf e}}
% \newcommand{\br}{{\bf r}}
% \newcommand{\bw}{{\bf w}}
% \newcommand{\de}{\stackrel{D}{=}}
% \newcommand{\bt}{\bigtriangleup}
% \newcommand{\bfequiv}{\mbox{\boldmath $\equiv$}}
% \newcommand{\bmu}{\mbox{\boldmath $\mu$}}
% \newcommand{\bnu}{\mbox{\boldmath $\nu$}}
% \newcommand{\bxi}{\mbox{\boldmath $\xi$}}
% \newcommand{\btau}{\mbox{\boldmath $\tau$}}
% \newcommand{\bgamma}{\mbox{\boldmath $\Gamma$}}
% \newcommand{\bphi}{\mbox{\boldmath $\Phi$}}
% \newcommand{\bfphi}{\mbox{\boldmath $\varphi$}}
% \newcommand{\bfeta}{\mbox{\boldmath $\eta$}}
% \newcommand{\bpi}{\mbox{\boldmath $\Pi$}}
% \newcommand{\bequiv}{\mbox{\boldmath $\equiv$}}
% \newcommand{\bvarepsilon}{\mbox{\boldmath $\varepsilon$}}
% \newcommand{\btriangle}{\mbox{\boldmath $\triangle$}}
% \newcommand{\bdelta}{\mbox{\boldmath $\Delta$}}
% \newcommand{\beps}{\mbox{\boldmath $\epsilon$}}
% \newcommand{\btheta}{\mbox{\boldmath $\theta$}}
% \newcommand{\balpha}{\mbox{\boldmath $\alpha$}}
% \newcommand{\bsphi}{\mbox{\boldmath $\varphi$}}
% \newcommand{\bsig}{\mbox{\boldmath $\sigma$}}
% \newcommand{\bfpsi}{\mbox{\boldmath $\psi$}}
% \newcommand{\bfdelta}{\mbox{\boldmath $\delta$}}
% \newcommand{\bsigma}{{\bf \Sigma}}
% \newcommand{\bzero}{{\bf 0}}
% \newcommand{\bpsi}{\mbox{\boldmath $\Psi$}}
% \newcommand{\bep}{\mbox{\boldmath $\epsilon$}}
% \newcommand{\bomega}{\mbox{\boldmath $\Omega$}}
% \newcommand{\bfomega}{\mbox{\boldmath $\omega$}}
% \newcommand{\blambda}{\mbox{\boldmath $\Lambda$}}
% \newcommand{\bflambda}{\mbox{\boldmath $\lambda$}}
% \newcommand{\bfsigma}{\mbox{\boldmath $\sigma$}}
% \newcommand{\bfpi}{{\mbox{\boldmath $\pi$}}}
% \newcommand{\bupsilon}{\mbox{\boldmath $\upsilon$}}
% \newcommand{\obs}{{\rm obs}}
% \newcommand{\mis}{{\rm mis}}
% \theoremstyle{plain}
% \newtheorem{criterion}{Criterion}
% \theoremstyle{definition}
% \newtheorem{condition}[theorem]{Condition}
% \usepackage[]{graphicx}
% \chardef\bslash=`\\ % p. 424, TeXbook
% \newcommand{\ntt}{\normalfont\ttfamily}
% \newcommand{\cn}[1]{{\protect\ntt\bslash#1}}
% \newcommand{\pkg}[1]{{\protect\ntt#1}}
% \let\fn\pkg
% \let\env\pkg
% \let\opt\pkg
% \hfuzz1pc % Don't bother to report overfull boxes if overage is < 1pc
% \newcommand{\envert}[1]{\left\lvert#1\right\rvert}
% \let\abs=\envert

@@

\begin{abstract}
Developing new imputation methodology has become a very active field. Unfortunately, there is no consensus on how to perform simulation studies to evaluate the properties of imputation methods. In this paper, we propose a move towards a standardized evaluation of imputation methodology. To demonstrate the need for standardization, we highlight a set of potential pitfalls that bring forth a chain of potential problems in the objective assessment of the performance of imputation routines. This may lead to sub-optimal use of  imputation in practice. Additionally, we suggest a course of action for simulating and evaluating missing data problems. [TODO: Narrowing the scope of the manuscript to the evaluation of methodology for statistical inference, not prediction or causal inference (although inferential validity encompasses predictive validity and may thus be generalized?). Emphasizing that the ideal evaluation of imputation methodology depends on the aim of the specific study (i.e. how the evaluated methods are used in practice). Split up the problems with the simulation setup/evaluation and the problems with the interpretation/extrapolation from these simulation studies. See e.g. \citet{gree17} on cognitive problem of generelization versus methodological problems. Explain how the standardisation leads to more neutrality, but that this is just a set of elements to consider/report, not a complete 'cookbook'. Add that we focus on  simulation studies evaluating imputation methodology for statistical inference.]
\end{abstract}

@@

Imputation is a state-of-the-art technique for drawing valid conclusions from incomplete data. The technique has earned a permanent spot in research and policymaking, demonstrated e.g. by the detailed manual created by the National Research Council \citep*{little2012prevention}. Although top-down enforcement of valid ways to handle missing data is not yet very pronounced, an increasing amount of researchers and data scientists are embracing imputation techniques. After all, the principle of imputation is very intuitive. The idea behind imputation is to impute (fill in) missing values, to obtain a valid estimate of what could have been. The completed data that are thus obtained can be analyzed by standard techniques, which conveniently splits up the process of solving the missing data problem and the complete-data problem. The quality of a solution obtained by imputation depends on the statistical properties of the incomplete data and the degree to which an imputation procedure is able to capture these properties when modeling missing values. In general, it holds that modeling missing data becomes more challenging when the amount of missingness increases. However, when (strong) relations in the data are present, the observed parts can hold great predictive power for the models that estimate the missingness. In that case, imputation would be substantially more efficient than the ubiquitous complete case analysis.

In contrast to ad hoc methods for dealing with missing values (e.g. list-wise deletion), imputation allows analysts to account for the sources of uncertainty that are related to the missingness problem. With the popular variant multiple imputation \citep{rubi76}, each missing value is replaced by a plausible data value several times, resulting in multiple completed data sets. The estimated parameters of interest for each of these completed sets can be combined into a single inference using Rubin's rules \citep[][p. 76]{rubi87}. [TODO: add that if the goal is statistical inference, the uncertainty due to missingness should be quantified/captured/propagated by the missing data method. Refer to \citet{sper20} for the context of prediction, where multiple imputation is not always necessary.] [TODO: paraphrase ``the quality of a technique can only be evaluated with respect to the aims of the problem/s it's intended to solve in practice".]

When evaluating the statistical properties (and thereby the practical applicability) of imputation methodology, researchers most often make use of simulation studies. Such studies are typically composed of an analysis pipeline in which data are generated, made incomplete, and imputed repeatedly under different simulation conditions, and a set of evaluation criteria is postulated to evaluate the performance of one or more missing data methods. However, there is no consensus on how to design, execute and report simulation studies aimed to evaluate imputation methodology. Without a `gold standard' for the evaluation of imputation routines, the validity and comparability of simulation setups may differ tremendously from one developer to another. This brings forth a chain of potential problems in the objective assessment of imputation method performance within and across studies, which may lead to sub-optimal use of imputation in practice.

% Especially with novel imputation methods being propagated from the fields of machine learning and artificial intelligence, the differences may become more pronounced. Although these promising new methods seem to yield even sharper imputations than now-standard imputation methods, the comparison may not be fair due to the simulation setup.

The purpose of this paper is threefold: First, to raise some concerns with respect to evaluating imputation methodology. These concerns stem from careful consideration with fellow `imputers' and from encounters as a reviewer for statistical journals. Second, to provide imputation methodologists with a suggested course of action when using simulation studies to evaluate imputation techniques for missing data problems. This suggested approach should identify common ground but is in no way intended as an absolute solution. This identifies the third purpose of this paper: discussion. We hope to elicit critical thinking regarding the problems at hand. We are all convinced that our methodology has some merit. But for sake of progress, it would be much more advantageous if the aim of our evaluations would go beyond \textit{proving the point} and would legitimately consider the statistical properties. [TODO: explain how the standardisation leads to more neutrality, but that this is just a set of elements to consider/report, not a complete 'cookbook'.]

Standardizing the evaluation of imputation methodology requires simulators to consider a couple of key aspects in their simulation workflows. Table \ref{table:steps} outlines some suggested steps to adopt. Please note that there is--of course--no `silver bullet' and that we do not claim to present a universally applicable approach. Our suggestions are the accumulated result of scientific literature, research experience, and discussion. For an excellent overview of general best practices for method evaluation by means of simulation, see \citet{morr18}. We recommend adhering to their proposed ADEMP structure (aims, data-generating mechanisms, estimands, methods, performance measures) for planning and reporting simulation studies, but highlight and add some specific recommendations for the evaluation of imputation methodology. [TODO: Add advice to pre-register simulation protocols, which allows for fair comparisons of per-protocol results \citep{pawe22}.]

@@

% [TODO: split up into design, execution and reporting? \citep{pawe22}]

The ideal evaluation of imputation methodology depends on the aim of the specific study (i.e. how the evaluated methods are used in practice). A simulation study developed for the comparison of imputation methods will typically have a different design than one aimed at establishing the inferential validity of a single (novel) method. The suitability of the evaluations should be assessed with the imputation methods' purpose in mind, since the simulation aims and setup are naturally intertwined with the choice of imputation method(s) under evaluation.

We limit the scope of this paper to comparative simulation studies in the context of statistical inference. We thereby exclude the evaluation of imputation methodology for prediction and causal inference. Simulation study designs aimed at comparing different methods' predictive performance in incomplete data typically do not start out from a complete data set in which missingness is induced by the simulator \citep{liu21}. Rather, the methods are evaluated against the observed values of a certain outcome variable (or target) as the estimand. After imputing the missingness in the incomplete predictor (or feature) space, a prediction method is applied to estimate the outcome. Pairs of imputation and prediction methods with high predictive accuracy in one or more benchmark data sets are deemed as `good'. Note that such a simulation design does not have a `ground truth', so only the comparative performance of methods may be established. We do not recommend this approach if the inferential validity of the imputations is of interest. For an overview of missingness in the prediction context see e.g. \citet{sper20}. Missing data within a causal inference framework has been described by e.g. \cite{more18} and \cite{moha21}.

Within the scope of comparative simulation studies for the evaluation of imputation methodology for statistical inference, we observe several problems. To demonstrate the broad impact of these problems, we subdivide the problems into the following four distinct categories: problems with simulation design, problems with data generation, problems with missingness generation, and problems with performance evaluation. We further detail the impact each of these problems may have on the validity of evaluations.

@@

\subsection{Simulation design}

% Before setting up a simulation study, the simulator should clearly define the scope of their evaluations. 
Estimand(s) or other simulation targets should be defined in the context of the study aim. [TODO: refer to \citet{pete14} about the need for a clear and unambiguous estimand.]

[TODO: describe problems with unclear simulation parameters, extrapolation beyond scope of simulations, QRPs \citep{pawe22} and misinterpretation \citep[see "investigator bias"][]{gree17}.]

[TODO: rewrite sampling variance paragraphs and combine with ``If there is more than one imputation method under evaluation, the simulation design should apply each method to every incomplete data set. Applying all methods to the same incomplete data set is computationally convenient, and minimizes unnecessary variation, which makes for fairer comparisons."]

[TODO: add nuance because this doesn't always hold, see Tims review!] Simulation studies on missing data methodology have the unique option to exclude sampling variance from their evaluations and only use the missing data generation procedure as the source of Monte Carlo variation. After all, we are interested in the missingness and are not considering the noise induced by the sampling mechanism for evaluation in such studies. Therefore, taking sampling variation into account is neither necessary for obtaining information about a method's ability to handle the missing data problem, nor for objectively comparing methods on their ability to correct for missingness \citep[see for a detailed discussion][]{vink14}. The only required change to simulation setups is to remove sampling variance from performance evaluation. Still, a lot of published work does not consider this option, while it could have sharpened inconclusive or obfuscated results. The comparative truth of the estimand(s) is determined by the choice of data-generating mechanism and whether sampling variance is included in the simulation setup, see Table \ref{table:dgm_truth}. If sampling variance cannot be omitted from the simulation scheme, multiple samples should be drawn from the data-generating mechanism (i.e., one sample per simulation repetition, a standard simulation setup). If sampling variance is not of interest, a single complete data set can be obtained from the data-generating mechanism (i.e., one sample for all simulation repetitions). This process is computationally convenient because only a single complete data set has to be considered during all of the simulations. We highly value the additional flexibility that such a simulation setup offers. Especially in the case of data transformations, it can be challenging to derive the true parametric references to evaluate against. Simply using a single generated complete set in which missingness is induced, and evaluating against that generated data set avoids a plethora of procedural problems and computational challenges.
When deviating from the standard simulation workflow, conventional pooling rules for multiple imputation \citep[cf.][p. 76-77]{rubi87} do not apply. Instead, alternative pooling rules need to be used \citep{raghunathan2003multiple,vink14}.

The simulation design should ideally be fully factorial, i.e. varying each simulation condition against all other conditions. There may, for example, be interactions between the source and amount of missingness in the incomplete data and the efficacy of imputation methods (i.e., the validity of the assumed missingness mechanism becomes increasingly important with higher missingness proportions). The number of simulation repetitions may be informed by the required level of precision in the simulation study \citep[e.g. as determined from a maximum tolerable level of uncertainty in terms of a performance measure's Monte Carlo error][]{morr18}.

@@

% [TODO: split up into data and missingness part of the DGM, with choices between data sources, missingness mechanisms and proportions. And divide problems between simulation and interpretation, see Reviewer 2: "whether method A (e.g. imputation) outperforms method B (e.g. listwise deletion) depends on the estimand, purpose of the study, used distributions (in the DGP, imputation model, possibly also analysis model), and simulation setup (e.g. dependency structure) - if a human misinterprets findings in a comparison of method A and B, then this is a cognitive problem and not the problem on how missingness was generated in the DGP."]

[TODO; describe problems with data generation and interpretation.]

- If the source of the data is model-based, the correlations in the data may be so high that there is no effective missingness due to the predictive power in the other variables, and the data source may be unfair if the data generating model is equal to one of the imputation models under evaluation.

- With a design-based simulation, the data source is unlikely to have no missingness and the choice of initial missing data method may propagate into the evaluations.

To evaluate the ability of an imputation routine to handle missingness, a form of truth has to be established. Those who perform simulation studies are in the luxury position to establish the truth beforehand by choosing a data-generating mechanism. Data-generating mechanisms define how a complete dataset is obtained at the start of each simulation repetition. There are two general approaches to generating complete data: (i) model-based simulation, in which data are drawn from a known statistical model or probability distribution, such as the multivariate normal distribution; and (ii) design-based simulation, where data are sampled (with replacement) from a sufficiently large observed set, such as official registers. The simulator should choose their data-generating mechanism(s) in line with the study scope. Model-based data-generating mechanisms have advantages in flexibility and precision, since data are generated from a known statistical model and the true theoretical parameters can be derived. A design-based approach is often used in situations where a probability distribution is not available, or where real-life data structures are of interest.

@@

@@

@@

[TODO: paraphrase ``be clear which mechanisms are of interest, how they are generated, if they are suitably generated".]

- Missingness mechanisms can be spurious (spurious MAR when correlations are too low, or when missingness proportion is too low). In reporting, the functional form or type of missingness is often omitted. As advice, MCAR can be a very informative mechnism for reference and MNAR may be useful for practice.

- Missingness proportions can be misinterpreted.

- Missingness pattern may have unexpected consequences (e.g., univariately generated multivariate missingness). In interpretation, watch out for generalizations from unintended special cases.

% Evaluating imputation methodology requires a missing data problem to be solved. After establishing a comparative truth from a data-generating mechanism, some form of missingness therefore has to be induced.

Often reports of simulation studies remain vague about the missingness conditions under investigation and, even worse, some authors only report something like:
\begin{quote}
\textit{We generated missing data following a missing at random missingness mechanism.}
\end{quote}
This should be considered unacceptable, as claims about the validity of the imputation inference heavily depend on the simulated missingness conditions, such as missingness mechanisms and missingness patterns. Missingness mechanisms describe the relationship between missing entries and observed data values whereas missingness patterns concern the location of missing entries across incomplete data \citep[][p. 8]{litt20}. Under this definition, there is a row-wise element to the missingness pattern describing which variables are jointly observed, and a column-wise element encompassing the amount of missingness in the data.

% \begin{table}[tb]
% \begin{center}
% \caption{Missingness mechanisms. [TODO: add refs \citet{seam13, meal15, dore18} for clarification and alternatives outlined in \citet{moha21, scho21, more18}]}
% \label{table:mech}
% \begin{tabular}{lll}
% \hline
%       & Interpretation and consequences of the mechanism \\
% \hline  
% MCAR  & Missing completely at random: The probability to be missing is the same for all \\
%       & cases. In other words, the missing values are missing at random and the observed \\
%       & values are observed at random. \\ \noalign{\smallskip}
% MAR   & Missing at random: The probability to be missing is the same within groups of \\
%       & cases defined by the observed data only. The essence is that observed relations in \\
%       & the data inform the missingness, such that MCAR may be assumed within the \\ 
%       & observed groups. \\ \noalign{\smallskip}
% v-MAR & Conditional on the fully observed variables missingness occurs at random 
%       & \citep[][p. 1026]{moha21}.\\
% MNAR  & Missing not at random: The probability to be missing depends on unobserved \\
%       & aspects of the data. The cause of the missingness is unknown and cannot be \\ 
%       & inferred from the observed data. This is considered non-ignorable missingness \\
%       & \citep[see e.g.][]{rubi76}. \\
% \hline
% \end{tabular}
% \end{center}
% \end{table}

Even the terminology on missingness generation can be confusing. Does a missingness proportion of 50\% mean that half of the entries in an incomplete dataset are missing, or that half of the rows have at least one missing entry? In this paper, we will henceforth refer to the latter as the proportion of incomplete cases, and keep the term `missingness proportion' restricted to the variable-by-variable interpretation. The distinction between the two concepts, however, is not always clear in the literature, and convoluting the terms may lead to incorrect generalizations because they rarely mean the same (e.g., a proportion of 50\% incomplete cases in bivariate data could translate to a missingness proportion of 25\% in both variables, or one completely observed variable and one with 50\% missingness).
